---
title: "Mapping Viet Nam with vnmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping Viet Nam with vnmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_sf <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  eval = has_sf
)
```

`vnmap` draws publication-ready choropleth maps of Viet Nam and accounts for
the July 2025 administrative reform, which replaced the 63 provincial-level
units with 34. It follows the compact API of
[`usmap`](https://usmap.dev).

```{r setup}
library(vnmap)
library(ggplot2)
```

## An outline map

`plot_vnmap()` with no data draws the current 34-unit outline.

```{r outline}
plot_vnmap(color = "white", linewidth = 0.2)
```

## Joining your own data

Supply a data frame, the column holding the values, and the column that
identifies each unit. Identifiers may be Vietnamese names (with or without
diacritics), English names, common aliases, or official codes.

```{r choropleth}
example <- data.frame(
  province = province_info()$name_en,
  value = seq_len(34)
)

plot_vnmap(
  data = example, values = "value", id = "province",
  color = "white", linewidth = 0.2
) +
  scale_fill_viridis_c()
```

The package bundles official 2024 indicators, so a GRDP-per-capita map is one
call away.

```{r grdp}
data(province_stats_2024)

plot_vnmap(
  data = province_stats_2024,
  values = "grdp_per_capita_million_vnd",
  id = "code",
  color = "white", linewidth = 0.2
) +
  scale_fill_viridis_c(name = "Million VND\nper person")
```

## Choosing a geography

The current 34-unit geography is the default. The 63-unit geography used
before 1 July 2025 remains available for historical data.

```{r geographies, eval = FALSE}
vn_map()                          # 34 current units as an sf object
vn_map("provinces_63")            # historical 63-unit geography
```

Always choose the geography that matches the reference period and coding
scheme of your data.

## Socio-economic regions

Every unit is tagged with one of the six General Statistics Office
socio-economic regions. Filter a map to a region with `region`, or look the
region up directly.

```{r regions}
plot_vnmap(region = "MRD", labels = TRUE)

province_region(c("HCMC", "Can Tho"))
head(province_info()[c("name_en", "region_code", "region_en")])
```

Assignments for the 63-unit geography are the official ones; assignments for
the merged 2025 units follow their dominant predecessor by population.

## Working without sf

The name, code, and region lookups, and the bundled statistics, do not require
the `sf` package. Only the geometry functions (`vn_map()`, `plot_vnmap()`,
`vnmap_crs()`) do. This chunk always runs.

```{r lookups, eval = TRUE}
library(vnmap)
province_code(c("Da Nang", "Danang", "HCMC"))
province_region("Bac Giang", geography = "provinces_63")
```
