---
title: "Mapping Viet Nam with vnmap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mapping Viet Nam with vnmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_sf <- requireNamespace("sf", quietly = TRUE) &&
  requireNamespace("ggplot2", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 6,
  eval = has_sf
)
```

`vnmap` draws publication-ready choropleth maps of Viet Nam and accounts for
the July 2025 administrative reform, which replaced the 63 provincial-level
units with 34. It follows the compact API of
[`usmap`](https://usmap.dev).

```{r setup}
library(vnmap)
library(ggplot2)
```

## An outline map

`plot_vnmap()` with no data draws the current 34-unit outline.

```{r outline}
plot_vnmap(color = "white", linewidth = 0.2)
```

## Joining your own data

Supply a data frame, the column holding the values, and the column that
identifies each unit. Identifiers may be Vietnamese names (with or without
diacritics), English names, common aliases, or official codes.

```{r choropleth}
example <- data.frame(
  province = province_info()$name_en,
  value = seq_len(34)
)

plot_vnmap(
  data = example, values = "value", id = "province",
  color = "white", linewidth = 0.2
) +
  scale_fill_viridis_c()
```

The package bundles official 2024 indicators, so a GRDP-per-capita map is one
call away.

```{r grdp}
data(province_stats_2024)

plot_vnmap(
  data = province_stats_2024,
  values = "grdp_per_capita_million_vnd",
  id = "code",
  color = "white", linewidth = 0.2
) +
  scale_fill_viridis_c(name = "Million VND\nper person")
```

## Choosing a geography

The current 34-unit geography is the default. The 63-unit geography used
before 1 July 2025 remains available for historical data.

```{r geographies, eval = FALSE}
vn_map()                          # 34 current units as an sf object
vn_map("provinces_63")            # historical 63-unit geography
```

Always choose the geography that matches the reference period and coding
scheme of your data.

## Current communes and the 2025 transition

The bundled commune layer is an observed 25 July 2026 snapshot. Its province
polygons are dissolved from the same source, so current parent and child
boundaries are spatially compatible. `valid_from` is deliberately missing:
the source snapshot does not establish a legal effective date for every code.

```{r communes, eval = FALSE}
hanoi <- vn_map("communes", province = "Ha Noi")
hanoi[c("code", "name_vi", "geography_vintage", "snapshot_date")]
```

The NSO July 2025 conversion workbook is not bundled because redistribution
terms have not been established. Downloading is explicit. The resulting table
is many-to-many and reports whether each 2025 target code is present in the
bundled 2026 snapshot; it contains no inferred area or population weights.

```{r crosswalk, eval = FALSE}
path <- download_administrative_crosswalk(tempfile(fileext = ".xlsx"))
cw <- administrative_crosswalk(path, province = "Ha Noi")
table(cw$relation, cw$current_code_status)
attr(cw, "current_snapshot_only_codes")
```

## Transport and logistics facilities

`infrastructure()` provides pinned redistributable aerodrome, port,
border-control and trunk-network geometry. These are community mapping rather
than an official register. Categories remain neutral, entity aliases and
border-control component roles are retained, and unknown lifecycle status is
preserved instead of treating every mapped feature as operational.

```{r infrastructure}
table(infrastructure(crs = 4326)$infrastructure_type)

plot_vnmap() +
  geom_infrastructure(type = c("aerodrome", "port"))
```

`aerodrome` does not establish commercial service and `port` does not establish
seaport status. Inspect `infrastructure_class`, `service_type`, `status_source`,
`facility_roles`, and `location_accuracy` before analysis. Province filtering
uses spatial intersection, so cross-province network segments are retained.
Network counts are OSM way segments rather than route counts. Lines lacking
the expected motorway, `CT`, `QL`, or railway tags are outside this snapshot's
selection and should not be interpreted as confirmed network absences.
For railways, filter `class = "main_line"` only when explicitly tagged
`usage=main`; yards, sidings, spurs, industrial and unspecified lines remain
separate. Network `province_code` is deliberately missing because ways can span
boundaries; `province_codes` and province filtering use polygon intersections.

## Socio-economic regions

Every unit is tagged with one of the six General Statistics Office
socio-economic regions. Filter a map to a region with `region`, or look the
region up directly.

```{r regions}
plot_vnmap(region = "MRD", labels = TRUE)

province_region(c("HCMC", "Can Tho"))
head(province_info()[c("name_en", "region_code", "region_en")])
```

Assignments for the 63-unit geography are the official ones; assignments for
the merged 2025 units follow their dominant predecessor by population.

## Industrial parks

`industrial_parks()` returns the national park registry as a provenance-rich
`sf` layer, one row per park. Use `province`, `status`, `category`, `accuracy`
or `include` to narrow it, and choose either the best available geometry,
polygons only, or representative points.

```{r industrial-parks}
head(industrial_parks(province = "Dong Nai", geometry = "point"))

plot_vnmap(include = "Dong Nai", color = "white", fill = "grey95") +
  geom_industrial_parks(province = "Dong Nai")
```

`industrial_park_registry()` returns the same records without geometry, which
does not need `sf` and is the natural table to join an external park list to,
either on `park_key` or on the administrative codes.

```{r industrial-park-registry}
registry <- industrial_park_registry(province = "Dong Nai")
head(registry[c("id", "name_vi", "commune_name", "area_ha", "location_accuracy")])
```

Records with `location_accuracy = "site"` carry a mapped boundary and an
`area_ha`. Records with `"locality"` are parks that OpenStreetMap names but has
not drawn; they are placed at the centre of the gates, stops, roads and plants
that name them, and are useful for locating a park but not for measuring it.

The registry includes only parks with a redistributable mapped location, so it
is smaller than the legal register and unevenly complete across provinces.
`industrial_park_coverage()` reports that gap, per province and against the
official national count.

```{r industrial-park-coverage}
head(industrial_park_coverage())
industrial_park_coverage("national")[c("metric", "value", "mapped", "coverage")]
```

Inspect `location_accuracy`, `geometry_source`, and `osm_refs` before using the
layer for site-level analysis; it is not a legal boundary register.

## Economic and policy zones

Economic zones are legally distinct from industrial parks. The conservative
first snapshot can be filtered by legal type, date, province, and status:

```{r economic-zones, eval=FALSE}
economic_zones(type = "border_gate_economic_zone", as_of = "2020-01-01")
plot_vnmap() + geom_economic_zones(type = "national_high_tech_park")
```

Rows retain legal instruments and official URLs where verified; missing
row-level instruments remain explicit candidate records. All initial
geometries are deterministic province-only points, not legal boundaries. The
current catalogue has 20 coastal and 26 border-gate records. The cited Ho Chi
Minh City annex names three EPZs; it is not a national completeness source. Unknown
establishment dates are retained by default in dated queries; use
`include_unknown = FALSE` for a strict panel. Economic zones can contain
industrial parks, but the two layers represent different legal entities.

## Working without sf

The name, code, and region lookups, and the bundled statistics, do not require
the `sf` package. Only the geometry functions (`vn_map()`, `plot_vnmap()`,
`vnmap_crs()`) do. This chunk always runs.

```{r lookups, eval = TRUE}
library(vnmap)
province_code(c("Da Nang", "Danang", "HCMC"))
province_region("Bac Giang", geography = "provinces_63")
```
